---
title: "eBird and Fire"
output: html_notebook
---

```{r}
library(dplyr)
library(ggplot2)
library(sf)
library(tidyverse)
library(anchors)
install.packages("anchors")
install.packages("RColorBrewer")
library(RColorBrewer)
```

# Split files using linux command line and read in first file of 100,000 lines (baa)
```{r}
system("less -n1 RawData/ebd_AU_relApr-2020.txt 2>&1", intern = FALSE, show.output.on.console = TRUE)
baa<- read.delim("RawData/baa.csv", header = TRUE, sep="\t")
head(baa)

```
# Filtering unwanted checklists
## Protocol used for filtering:  
Removed Incomplete checklists
Removed checklists Duration.Minutes less than 10m and more than 5hrs
Removed Checklists Effort.Distance more than 10km  
# Selecting wanted columns
Keeping Locality ID, Sampling event identifier, Lat and Long, Date
# Selecting distinct rows
Removing duplicates 


```{r}
baatrimmed<- baa %>% 
  filter(ALL.SPECIES.REPORTED==1, EFFORT.DISTANCE.KM<10, DURATION.MINUTES<300, DURATION.MINUTES>10) %>% 
  select(SAMPLING.EVENT.IDENTIFIER, LOCALITY.ID, LATITUDE, LONGITUDE, OBSERVATION.DATE) %>% 
  distinct()
```

# Spacial Processing steps:

## Read in fire shapefile
```{r}
fireshape<- st_read("National_Indicative_Aggregated_Fire_Extent_Dataset_v20200211/National_Indicative_Aggregated_Fire_Extent_Dataset_v20200211.shp")
```


## Map of shapefile
```{r}
ggplot() + 
  geom_sf(data = fireshape, size = 0.1, color = "black", fill = "black") + 
  ggtitle("Fire Boundary Plot") + 
  coord_sf()
```
## Convert ebird checklists to sf object with CRS code from fire shapefile
Make the eBird data a sf object, then match CRS(Coordinate Reference System) with fire shapefile
```{r}
ebird_points <- st_as_sf(baatrimmed, coords=c("LONGITUDE", "LATITUDE"), crs=st_crs(fireshape))

```
## Make a map of the points using ggplot and geom_sf on the fire shapefile map
```{r}
ggplot() + 
  geom_sf(data = fireshape, size = 0.1, color = "black", fill = "black") + 
  geom_sf(data=ebird_points, size=0.5, color="green")+
  ggtitle("Fire Boundary with listpoints") + 
  coord_sf()
```
## add column for inside/outside using st_join
```{r}
pointsinout<-st_join(ebird_points, fireshape, join = st_within)
View(pointsinout)
count(pointsinout, Id)
list(pointsinout$Id)

inout<- replace.value(pointsinout, "Id", from = 0, to = 1)
inout[is.na(inout)] = 0
count(inout, Id)
inout1<- replace.value(inout, "Id", from = 0, to = 2)
```


## make a map with points coloured by inside/outside
```{r}
ggplot() + 
  geom_sf(data = fireshape, size = 0.1, color = "blue", fill = "blue") + 
  geom_sf(data=inout1, size=0.5, colour = inout1$Id)+
  ggtitle("Points by In-Out") + 
  coord_sf()
  
display.brewer.pal(n = 2, name = "Set2")
```

# write out as rds to a directory called proceessed_data
```{r}

```