---
title: "eBird and Fire"
output: html_notebook
---



```{r}
system("less -n1 RawData/ebd_AU_relApr-2020.txt 2>&1", intern = FALSE, show.output.on.console = TRUE)
baa<- read.delim("RawData/baa.csv", header = TRUE, sep="\t")
head(baa)

```
# Filtering unwanted checklists
## Protocol used for filtering:  
Removed Incomplete checklists
Removed checklists Duration.Minutes less than 10m and more than 5hrs
Removed Checklists Effort.Distance more than 10km  
# Selecting wanted columns
## Protocol:  
Keeping Locality ID, Sampling event identifier  
# Selecting distinct rows
## Protocol
Removing duplicates 
```{r}
library(dplyr)
baatrimmed<- baa %>% 
  filter(ALL.SPECIES.REPORTED==1, EFFORT.DISTANCE.KM<10, DURATION.MINUTES<300, DURATION.MINUTES>10) %>% 
  select(SAMPLING.EVENT.IDENTIFIER, LOCALITY.ID) %>% 
  distinct()
```

```{r}

```

# Spacial Processing steps:

# Read in fire shapefile
```{r}

```
# Map of shapefile
```{r}

```
# Convert ebird checklists to sf object with CRS code from fire shapefile
```{r}

```
# Make a map of the points using ggplot and geom_sf on the fire shapefile map
```{r}

```
# add column for inside/outside using st_join
```{r}

```
# make a map with points coloured by inside/outside
```{r}

```

# write out as rds to a directory called proceessed_data
```{r}

```