---
title: "Josh Lee's BEES3041 Code"
output: html_notebook
---

# Libaries used:
```{r}
library(dplyr)
library(ggplot2)
library(sf)
library(tidyverse)
library(anchors)
library(readr)
library(tidyverse)
library(lubridate)
library(rnaturalearth)
library(mgcv)

```

# Split files using a linux command line and read in first file (baa)
Since the data from eBird was so large, we split into 230 files with 100,000 rows
```{r}
system("less -n1 RawData/ebd_AU_relApr-2020.txt 2>&1", intern = FALSE, show.output.on.console = TRUE)
baa<- read.delim("Data/Raw/RawEbird/baa", header = TRUE, sep="\t")
head(baa)
view(baa)
```
# Filtering unwanted checklists
## Protocol used for filtering and selecting:  
Removed Incomplete checklists  
Removed checklists Duration.Minutes less than 10m and more than 5hrs  
Removed Checklists Effort.Distance more than 10km  
Keeping Locality ID, Sampling event identifier, Lat and Long, Date  
Removing duplicates 

```{r}
baatrimmed<- baa %>% 
  filter(ALL.SPECIES.REPORTED==1, EFFORT.DISTANCE.KM<10, DURATION.MINUTES<300, DURATION.MINUTES>10) %>% 
  dplyr::select(SAMPLING.EVENT.IDENTIFIER, LOCALITY.ID, LATITUDE, LONGITUDE, OBSERVATION.DATE) %>% 
  distinct()
```

# Spacial Processing steps:
## Read in fire shapefile
National Indicative Aggregated Fire Extent Datasets from Department of Agriculture Water and the Environment  
```{r}
fireshape<- st_read("Data/Raw/National_Indicative_Aggregated_Fire_Extent_Dataset_v20200211/National_Indicative_Aggregated_Fire_Extent_Dataset_v20200211.shp")
```

## Map of shapefile
```{r}
ggplot() + 
  geom_sf(data = fireshape, size = 0.1, color = "black", fill = "black") + 
  ggtitle("Fire Boundary Plot") + 
  coord_sf()
```
## Convert ebird checklists to sf object with CRS code from fire shapefile
Make the eBird data a sf object, then match CRS(Coordinate Reference System) with fire shapefile
```{r}
ebird_points <- st_as_sf(baatrimmed, coords=c("LONGITUDE", "LATITUDE"), crs=st_crs(fireshape))
```
## Make a map of the points using ggplot and geom_sf on the fire shapefile map
```{r}
ggplot() + 
  geom_sf(data = fireshape, size = 0.1, color = "black", fill = "black") + 
  geom_sf(data=ebird_points, size=0.5, color="green")+
  ggtitle("Fire Boundary with listpoints") + 
  coord_sf()
```
## Add column for inside/outside using st_join
Plotting didnt like the NAs or 0s so had to convert to 1s(Inside burnt zone) and 2s(Outside burnt zone)

```{r}
pointsinout<-st_join(ebird_points, fireshape, join = st_within)
View(pointsinout)
count(pointsinout, Id)
list(pointsinout$Id)

inout<- replace.value(pointsinout, "Id", from = 0, to = 1)
inout[is.na(inout)] = 0
inout1<- replace.value(inout, "Id", from = 0, to = 2)
count(inout, Id)
```

## Make a map with points coloured by inside/outside
Blue=Fire extent, Red=Checklists outisde, Black=Checklists Inside
```{r}
MapbyInOut<- ggplot() + 
  geom_sf(data = fireshape, size = 0.1, color = "blue", fill = "blue") + 
  geom_sf(data=inout1, size=0.5, colour = inout1$Id)+
  ggtitle("Points by In-Out") + 
  coord_sf()
print(MapbyInOut)
```

# Write out as rds to a directory called processed_data
```{r}
saveRDS(MapbyInOut, "Data/Processed/MapbyInOut.rds")
```

# Get a column names vector for usage in the next function
```{r}
fireshape_current<- st_read("Data/Raw/National_Indicative_Aggregated_Fire_Extent_Dataset_v20200211/National_Indicative_Aggregated_Fire_Extent_Dataset_v20200211.shp")


baa <- read_delim("Data/Raw/RawEbird/baa", "\t", 
                  escape_double = FALSE, trim_ws = TRUE)
colnames(baa)<-gsub(" ",".",colnames(baa))

column_names <- colnames(baa)
```

# Writing the function to read in, filter data and convert to spatial feature matched with fire shapefile
```{r}
read_and_process_raw_ebird <- function(filename,fireshape=fireshape){
  #system("less -n1 RawData/ebd_AU_relApr-2020.txt 2>&1", intern = FALSE, show.output.on.console = TRUE)
  baa <- read_delim(filename, "\t", 
                    escape_double = FALSE, trim_ws = TRUE,
                    col_names = FALSE)
  
  baa <- if (baa[1,1]=="GLOBAL UNIQUE IDENTIFIER") {
    baa %>% 
      slice(2:100000)
  } else {
    baa
  }
  
  colnames(baa) <- column_names
  
  
  baatrimmed <- baa %>% 
    filter(ALL.SPECIES.REPORTED==1, EFFORT.DISTANCE.KM<10, DURATION.MINUTES<300, DURATION.MINUTES>10) %>% 
    dplyr::select(SAMPLING.EVENT.IDENTIFIER, LOCALITY.ID, LATITUDE, LONGITUDE, OBSERVATION.DATE) %>% 
    distinct()
  
  ebird_points <- st_as_sf(baatrimmed, coords=c("LONGITUDE", "LATITUDE"), crs=st_crs(fireshape))
  
  pointsin <- fireshape %>% st_join(ebird_points) %>% 
    st_set_geometry(NULL) %>% 
    dplyr::select(SAMPLING.EVENT.IDENTIFIER) %>% 
    mutate(in_fire="True")
  
  pointsinout <- baatrimmed %>%
    left_join(., pointsin)
  
  saveRDS(pointsinout,file = paste0("Data/Processed/filteredData/", gsub("RawData/", "",filename), ".rds"))
}
```

# Apply function to all 230 files
```{r}
all_files<-list.files("Data/Raw/RawEbird/",full.names = TRUE)


for(i in 1:length(all_files)){
  read_and_process_raw_ebird(filename=all_files[i],fireshape=fireshape_current)
}
```

# Bind outputs of read_and_process_raw_ebird to one data frame
```{r}
all_files<-list.files("Data/Processed/filteredData",full.names = TRUE)
first_list<-list()


for (i in 1:length(all_files)){
  print(i)
  temp<-readRDS(all_files[i])
  temp$LATITUDE<-as.numeric(temp$LATITUDE)
  temp$LONGITUDE<-as.numeric(temp$LONGITUDE)
  temp$OBSERVATION.DATE<-as.character(temp$OBSERVATION.DATE)
  first_list<-bind_rows(first_list,temp)
}
```

# Counting data in relation to fire
```{r}
first_list$in_fire <- recode(first_list$in_fire,True = "In Fire")
first_list$in_fire <- replace_na(first_list$in_fire,"Not in Fire")

table(first_list$in_fire)

first_list$OBSERVATION.DATE<-ymd(first_list$OBSERVATION.DATE)
first_list$after_fire<-first_list$OBSERVATION.DATE > ymd("2020-01-10")

table(first_list$in_fire,first_list$after_fire)
```

# Filtering out checklists outside of fire, define column names using first file
```{r}
list_infire<- filter(first_list, in_fire == "In Fire")
baa <- read_delim("Data/Raw/RawEbird/baa", "\t", escape_double = FALSE, trim_ws = TRUE)
colnames(baa)<-gsub(" ",".",colnames(baa))
column_names <- colnames(baa)
baa_inside<- semi_join(baa, list_infire, by = "SAMPLING.EVENT.IDENTIFIER")
```

# Writing function to extract species data for these checklists in all data files
```{r}
extract_species_data <- function(filename, list_infire){
  baa <- read_delim(filename, "\t", 
                    escape_double = FALSE, trim_ws = TRUE,
                    col_names = FALSE)
  
  baa <- if (baa[1,1]=="GLOBAL.UNIQUE.IDENTIFIER") {
    baa %>% 
      slice(2:100000)
  } else {
    baa
  }
  
  colnames(baa) <- column_names
  
  
  baa_inside<- semi_join(baa, list_infire, by = "SAMPLING.EVENT.IDENTIFIER")
  
  
  saveRDS(baa_inside,file = paste0("Data/Raw/rawspeciesdata/", gsub("Data/Raw/RawEbird/", "",filename), ".rds"))
}


all_files<-list.files("RawData",full.names = TRUE)


for(i in 1:length(all_files)){
  extract_species_data(filename=all_files[i], list_infire)
}
```

# Binding these output files into one data frame with all bird observations from within fire boundary 

```{r}
all_files<-list.files("Data/Raw/rawspeciesdata",full.names = TRUE)
second_list<- list()

for (i in 1:length(all_files)){
  print(i)
  temp<-readRDS(all_files[i])
   second_list<-bind_rows(second_list,temp)
}
```

# Change working directory to folder just created, read in all files and define date variable using a fuction, also making minor string changes 
```{r}
setwd("Data/Raw/rawspeciesdata/")

read_dat_function <- function(file_name) {
  dat <- readRDS(file_name)
  dat$OBSERVATION.DATE<-ymd(dat$OBSERVATION.DATE)
  return(dat)
}

files <- list.files("../rawspeciesdata/")

data <- lapply(files, read_dat_function)
data_df <- do.call(rbind, data)
data_df$LATITUDE<-as.numeric(data_df$LATITUDE)
data_df<-filter(data_df,LATITUDE < -25)

length(unique(data_df$SAMPLING.EVENT.IDENTIFIER))
length(unique(data_df$COMMON.NAME))
```

# Reset wd, write another function to locate and assign checklists with species x and checklists without species x. Apply cutoff date for before/after fire event (10th Jan 2020). Create GAM for species x and write into new folder (model_objects)
```{r}
setwd("..")

process<-function(species_name,data_df=data_df){
  sl <- data_df %>%
    dplyr::filter(COMMON.NAME == species_name) %>%
    mutate(present=1)
  
  sl_lists <- sl %>%
    dplyr::select(SAMPLING.EVENT.IDENTIFIER) %>%
    distinct()
  
  lists_without <- data_df %>%
    dplyr::filter(! SAMPLING.EVENT.IDENTIFIER %in% sl_lists$SAMPLING.EVENT.IDENTIFIER) %>%
    dplyr::select(SAMPLING.EVENT.IDENTIFIER,OBSERVATION.DATE, DURATION.MINUTES, LATITUDE, LONGITUDE,
                  EFFORT.DISTANCE.KM, COUNTY.CODE, OBSERVER.ID) %>% 
    distinct() %>%
    mutate(present=0)
  
  final_sl_dat <- sl %>%
    bind_rows(lists_without) %>%
    mutate(DURATION.MINUTES=as.numeric(DURATION.MINUTES),
           EFFORT.DISTANCE.KM=as.numeric(EFFORT.DISTANCE.KM)) %>%
    mutate(MONTH=month(OBSERVATION.DATE)) %>%
    mutate(before.after=ifelse(OBSERVATION.DATE>ymd("2020-01-10"), "After", "Before"))
  
  mod <- mgcv::gam(present ~ before.after + s(DURATION.MINUTES) +
                     s(EFFORT.DISTANCE.KM) + s(MONTH, bs="cc", k=11), 
                   family="binomial", data=final_sl_dat)
  
  saveRDS(mod, paste0("Output/data/model_objects/", gsub(" ", "_", species_name), ".RDS"))
  
  final_sl_dat %>%
    group_by(before.after) %>%
    summarize(percent_observed=sum(present)/n())->out
  return(out)
}
```

# Apply this function to all species in the df with more than 500 observations 
## Included funny trick to keep species names
```{r}
data_df %>%
  group_by(COMMON.NAME) %>%
  count(num.obs=n()) %>%
  filter(num.obs>500)->a

setNames(as.list(a$COMMON.NAME),a$COMMON.NAME) %>%
  map_df(process,data_df,.id="var") -> out

drop<- out %>%
  spread(key="before.after",value="percent_observed") %>%
  mutate(percentage_drop=(`After`-`Before`)) %>%
  arrange(percentage_drop)

```

# Writing function to extract model summaries and map_df into one df

```{r}
all_files <- list.files("Output/data/model_objects/")

  extractingcoeffse_function <- function(file_name) {
    data <- readRDS(paste0("Output/data/model_objects/", file_name))
    datsum<- summary(data)
    se<- datsum$se
    deviance<- datsum$dev.expl
    n<-datsum$n
    wgoods<- as.data.frame(summary(data)$p.tab) %>%
      rownames_to_column(var="Term") %>%
      mutate(deviance=deviance) %>%
      mutate(N=n) %>%
      mutate(converged=data$converged) %>%
      mutate(species=gsub(".RDS", "", file_name))
    return(wgoods)
  }
  
  bird_response_df <- map_df(all_files, extractingcoeffse_function)
```

# removing waterbirds
```{r}
bird_respones_new <- dplyr::filter(bird_response_df, !(bird_response_df$species == "Australian_Pelican" | 
                                                       bird_response_df$species == "Chestnut_Teal" |
                                                       bird_response_df$species == "Great_Cormorant"|
                                                       bird_response_df$species == "Little_Black_Cormorant"|
                                                       bird_response_df$species == "Little_Pied_Cormorant"|
                                                       bird_response_df$species == "Maned_Duck"|
                                                       bird_response_df$species == "Pacific_Black_Duck"|
                                                       bird_response_df$species == "Silver_Gull"|
                                                       bird_response_df$species == "White-faced_Heron"))

```

# Read in trait data obtained from (Garnett et al., 2015)
T Garnett, Stephen; E Duursma, Daisy; Ehmke, Glenn; Guay, Patrick-Jean; Stewart, Alistair; K Szabo, Judit; et al. (2015): Australian Bird Data Version 1.0 (2015). figshare. Collection. https://doi.org/10.6084/m9.figshare.1499292

```{r}
traits<- read.csv("Data/Raw/trait_data.csv")
```

# Select only traits of interest, and renaming species where there were discrepancies between the two dfs
```{r}
fire_coefs<- rename(bird_respones_new, species_name = species)


foodtraits<- traits %>% 
  dplyr::select("X3_Taxon_common_name_2", "X99_Body_mass_average_8", "X163_Food_Fruit_10":"X173_Food_fish_or_invertebrates_Inland_waters_10", "X115_Feeding_habitat_Terrestrial_Arid_shrubland_9":"X145_Feeding_habitat_Urban_landscapes_9", "X193_National_movement_local_dispersal_13":"X197_National_movement_Irruptive_13", "X126_Feeding_habitat_Inland_waters_Rivers_and_streams_9", "X127_Feeding_habitat_Inland_waters_Deep_open_water_9":"X135_Feeding_habitat_Coastal_Mangrove_floor_9") %>% 
  rename(species_name = "X3_Taxon_common_name_2") %>% 
  mutate(species_name=gsub(" ", "_", species_name)) 


foodtraits$species_name<- recode(foodtraits$species_name, 'Black-faced_Cuckoo-shrike' = "Black-faced_Cuckooshrike",
                   'Common_Blackbird' = "Eurasian_Blackbird",
                   'Common_Starling' = "European_Starling",
                   'Grey_Butcherbird' = "Gray_Butcherbird",
                   'Grey_Fantail' = "Gray_Fantail",
                   'Grey_Shrike-thrush' = "Gray_Shrikethrush",
                   'Australian_Wood_Duck' = "Maned_Duck",
                   'Red-browed_Finch' = "Red-browed_Firetail",
                   'Scarlet_Honeyeater' = "Scarlet_Myzomela",
                   'Superb_Fairy-wren' = "Superb_Fairywren",
                   'Variegated_Fairy-wren' = "Variegated_Fairywren",
                   'Willie_Wagtail' = "Willie-wagtail", 
                   'Jacky_Winter' = "Jacky-winter")

traitsfin <- foodtraits %>% 
  left_join(fire_coefs, foodtraits, by = "species_name") %>% 
  filter(species_name %in% fire_coefs$species_name)
```

## Minor string changes 
```{r}
values<- traitsfin %>% 
  filter(Term == "before.afterBefore") %>% 
  mutate(X99_Body_mass_average_8 = as.numeric(X99_Body_mass_average_8))

values$X193_National_movement_local_dispersal_13 <- as.integer(values$X193_National_movement_local_dispersal_13)
values$X194_National_movement_Partial_migrant_13 <- as.integer(values$X194_National_movement_Partial_migrant_13)
values$X195_National_movement_Total_migrant_13 <- as.integer(values$X195_National_movement_Total_migrant_13)
values$X196_National_movement_Nomadic_or_opportunistic_13 <- as.integer(values$X196_National_movement_Nomadic_or_opportunistic_13)
values$X197_National_movement_Irruptive_13 <- as.integer(values$X197_National_movement_Irruptive_13)

```

# Calculation of habitat specialism and feeding specialism
## These use original variables from the trait data to make new variables that will be used in our linear models.
```{r}
valuesfin<- values %>% 
  mutate(foodspec = X163_Food_Fruit_10 +                                                          
         X164_Food_Nectar_or_pollen_10 +                                                 
         X165_Food_Seeds_10 +                                                              
         X166_Food_Foliage_or_herbs_10 +                                                   
         X167_Food_Corms_or_tubers +                                                       
         X168_Food_Terrestrial_invertebrates_10 +                                          
         X169_Food_Terrestrial_vertebrates_10 +                                            
         X170_Food_Carrion_10 +                                                            
         X171_Food_Intertidal_invertebrates_10 +                                           
         X172_Food_Fish_or_invertebrates_Marine_10 +                                       
         X173_Food_fish_or_invertebrates_Inland_waters_10) %>% 
  mutate(habitatspec = X115_Feeding_habitat_Terrestrial_Arid_shrubland_9 +
           X116_Feeding_habitat_Terrestrial_Chenopod_shrubland_9 +
           X117_Feeding_habitat_Terrestrial_Heath_9 + 
           X118_Feeding_habitat_Terrestrial_Triodia_hummock_grassland_9 + 
           X119_Feeding_habitat_Terrestrial_Other_grassland_9 + 
           X120_Feeding_habitat_Terrestrial_Mallee_9 + 
           X121_Feeding_habitat_Terrestrial_Tropical_savanna_woodland_9 + 
           X122_Feeding_habitat_Terrestrial_Temperate_dry_sclerophyll_forest_and_woodland_9 +
X123_Feeding_habitat_Terrestrial_Temperate_wet_sclerophyll_forest_and_woodland_9 +
X124_Feeding_habitat_Terrestrial_Rainforest_9 +                                   
X125_Feeding_habitat_Terrestrial_Mangrove_trees_9 +                               
X126_Feeding_habitat_Inland_waters_Rivers_and_streams_9 +                         
X127_Feeding_habitat_Inland_waters_Deep_open_water_9 +                            
X128_Feeding_habitat_Inland_waters_Shallow_open_water_9 +                         
X129_Feeding_habitat_Inland_waters_Reeds_and_tall_wet_grassland_9 +               
X130_Feeding_habitat_Inland_waters_Low_marshland_and_wet_grassland_9 +            
X131_Feeding_habitat_Coastal_Sandy_9 +                                            
X132_Feeding_habitat_Coastal_Rocky_9 +                                            
X133_Feeding_habitat_Coastal_Soft_mud_9 +                                         
X134_Feeding_habitat_Coastal_Saltmarsh_9 +                                        
X135_Feeding_habitat_Coastal_Mangrove_floor_9 +                                   
X136_Feeding_habitat_Marine_Very_cold_pelagic_9 +                                 
X137_Feeding_habitat_Marine_Cold_pelagic_9 +                                      
X138_Feeding_habitat_Marine_Temperate_pelagic_9 +                                 
X139_Feeding_habitat_Marine_Warm_pelagic_9 +                                      
X140_Feeding_habitat_Marine_Cold_inshore_9 +                                      
X141_Feeding_habitat_Marine_Temperate_inshore_9 +                                 
X142_Feeding_habitat_Marine_Warm_inshore_9 +                                      
X143_Feeding_habitat_Other_non.Australian_habitat_9 +                             
X144_Feeding_habitat_Agricultural_landscapes_9 +
X145_Feeding_habitat_Urban_landscapes_9) %>% 
  mutate()
```

# Running linear models between fire response and food specialism, habitat specialism, size and sedentism.
## Place results into final data frame
```{r}
#### models
modfood<-lm(Estimate ~ foodspec, data = valuesfin,weights = 1/`Std. Error`)
modhabitat<-lm(Estimate ~ habitatspec, data = valuesfin,weights = 1/`Std. Error`)
modsize<-lm(Estimate ~ X99_Body_mass_average_8, data = valuesfin,weights = 1/`Std. Error`)
modmobility<-lm(Estimate ~ X193_National_movement_local_dispersal_13, data = valuesfin,weights = 1/`Std. Error`)

summary(modfood)
summary(modhabitat)
summary(modsize)
summary(modmobility)

food <- broom::glance(modfood)
habitat<- broom::glance(modhabitat)
size<- broom::glance(modsize)
sedentism<- broom::glance(modmobility)

outputs<- bind_rows(habitat, food, size, sedentism)
outputs$test <- c("Habitat Specialism", "Food Specialism", "Body Size", "Sedentism")

write.csv(outputs, "Output/results.csv")
results <- read.csv("Output/results.csv")
```

# Plotting models to visualise results
```{r}

df<- valuesfin %>% 
  mutate(Estimate, value = Estimate*-1)

ggplot(modfood, aes(foodspec, Estimate*-1)) + geom_point() + theme_bw() + xlab("Number of Feeding Guilds") + ylab("Fire Response")
ggsave("Output/figures/foodmodel.PNG")
ggplot(modsize, aes(X99_Body_mass_average_8, Estimate*-1)) + geom_point()+ scale_x_log10()+ theme_bw() + xlab("Body Mass") + ylab("Fire Response")
ggsave("Output/figures/sizemodel.PNG")
ggplot(modhabitat, aes(habitatspec, Estimate*-1)) + geom_point()+ theme_bw() + xlab("Number of Habitats") + ylab("Fire Response")
ggsave("Output/figures/habitatmodel.PNG")
ggplot(modmobility, aes(X193_National_movement_local_dispersal_13, Estimate*-1)) + geom_point()+ theme_bw() + xlab("Sedentism") + ylab("Fire Response")
ggsave("Output/figures/mobilitymodel.PNG")

```

